---
title: "1b_TidyverseDataWrangling"
author: "Aaron C Cochran & Emilie Henderson"
date: "February 13, 2018"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Tidyverse Data Manipulation

This section of the workshop will cover the specific features related to data manipulation using the `tidyverse` collection of packages. It isn't intended to replace Base R but offers a complimentary skillset that is being widely used in the R universe as `tidyverse` has been written over the past few years. 

# The pipeline operator

` %>% ` is known as the pipeline operator. `x %>% y` is functionally equivalent to `f(y(x))`: `y(x)` serves as the input to `f()`, and `x` is the input for `y()`. It says "take the object on the left side of the pipeline and do the thing on the right side of the pipeline to that object."

# `dplyr` verbs

* `filter` - Keeps data based on filtered criteria.
* `select` - Selects columns of the data.frame
* `arrange` - 
* `mutate` - Create new variables from other columns
* `summarise` - Aggregate and create summary variables like `mean`, `sd` and counts. 
* `group_by` - Define groups based on variables, then call `summarise` to aggregate over groups. 

`dplyr` also introduces functions that will be recognizable to anyone working with SQL. 

* `left_join` - return all rows from x, and all columns from x and y. Rows in x with no match in y will have NA values in the new columns. If there are multiple matches between x and y, all combinations of the matches are returned.
* `anti_join` - return all rows from x where there are not matching values in y, keeping just columns from x.
* `semi_join` - return all rows from x where there are matching values in y, keeping just columns from x.

```{r}
?dplyr::left_join
```

## In practice

```{r}
library(nycflights13)
library(tidyverse)
data(flights)
flights
```

### `filter` rows

base R method for comparison

```{r, eval=FALSE}
flights[flights$month==4, ]
```




```{r}
flights %>% filter(month  == 4) %>% head(10)# select April flights only, return first 10
```

```{r}
flights %>% filter(month == 4 & day == 23) %>% head(10) # combined filter of month and day, return first 10
```

```{r}
flights %>% filter(month == 4 & day == 23 | month == 4 & day == 24) # filtering all flights on April 23 or 24th. 

flights %>% filter(month == 4 & day ==23:24) # much less verbose

flights %>% filter(month == 4 & day == c(1, 4, 6) & carrier == "UA" & arr_time <= 913) # use c() when you need to not use sequential numbers
```


### `select` columns

```{r}
flights %>% filter(month == 4 & day == 1 & carrier == "UA") %>% select(-year, -month, -day, -carrier) # use - for "not"
```

### `arrange` rows

```{r}
# base R method

flights[order(flights$dep_delay), c("carrier", "dep_delay")]
```

```{r}
flights %>%
  select(carrier, dep_delay) %>%
  arrange(dep_delay)
```

### `mutate`: creating new variables

Let's approximate speed in mph using $speed = \dfrac{dist}{time}$ 

```{r}
# base R approach
flights$speed <- flights$distance / flights$air_time*60
flights[, c("distance", "air_time", "speed")]
```


```{r}
flights %>%  mutate(speed = distance/air_time*60) %>% select(distance, air_time, speed)
```


# `summarise` 

> Note: UK or US English spellings are acceptable. 

`summarise` (or `summerize`) 

```{r}
# summary stats in tidy format
flights %>% 
  mutate(speed = distance/air_time*60) %>% 
  select(distance, air_time, speed) %>% 
  summarise(avg_speed = mean(speed))


# why is there an NA for average speed?  (hint: type mean(c(1, 2, NA))
flights %>%  
  mutate(speed = distance/air_time*60) %>% 
  select(distance, air_time, speed) %>% 
  summarise(avg_speed = mean(speed, na.rm=TRUE))

```

For functions like `mean` we need to include `na.rm=TRUE` to specify that all missing values (`NA`) are removed before the function calculates the result. 


Other aggregate functions:

* `min(x)` - min value of `x`
* `max(x)` - max value of `x`
* `mean(x)` - mean value of `x`
* `median(x)` - median value of `x`
* `quantile(x, p)` - `p`th quantile of vector `x`
* `sd(x)` - standard deviation of vector `x`
* `var(x)` - variance of vector `x`
* `IQR(x)` - Inter Quartile Range (IQR) of vector `x`
* `diff(range(x))` - total range of the vector `x`


