---
title: "Data visualization"
author: "Aaron C Cochran & Emilie Henderson"
date: "February 16, 2018"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE)
library(tidyverse)
library(plotly)
```

# Data visualization in R and tidyverse

## base R way

```{r}
# I don't know much about base R plotting, so if you want this section please take it
```


## `ggplot2`

In the `tidyverse` collection there is a package called `ggplot2`. It provides a "grammar of graphics" approach to plotting, where you create plots layer by layer and arrange the plots as you see fit. 



```{r, eval=FALSE}

ggplot(data = MYDATA, aes(X=MY_X, Y=MY_Y, color=MY_COLOR)) +
  geom_()  # Many types of geom in ggplot2. 
# If you use the development version, you can even do simple features (spatial)

```


```{r}
library(ggplot2)
library(nycflights13)

data(flights)

ggplot(data = flights, aes(x=month)) +
  geom_bar()

```

```{r}
# better, cleaned up version
ggplot(flights, aes(factor(month))) + geom_bar() + ggtitle('NY Flights by Month, 2013') + xlab('Month') + ylab('Count') + theme_minimal()
```

### `fill` and `color` 

```{r}
p1 <- ggplot(flights, aes(factor(month))) 

p1 + geom_bar(aes(fill=carrier))

```

```{r}
ggplot(filter(flights, carrier == c('AA', 'DL', 'UA'))) + geom_point(aes(x=dep_delay, y=arr_delay, color=carrier))
```

### Using `dplyr` to arrange data before plotting

The `carrier` field in our flights data uses carrier codes to identify each carrier. What if we wanted the actual names for the carriers so we could make a publication-quality plot?

```{r}
data(airlines)
airlines

data(flights)
flights
```

```{r}
flights %>% select(carrier, dep_delay, arr_delay) %>% left_join(airlines, by="carrier") %>% filter(carrier == c("AA","UA", "DL")) %>% 
  ggplot() + geom_point(aes(x=dep_delay, y=arr_delay, color=name), alpha=.5) +
  ggtitle('Arrival and departure delays, by Airline') +
  xlab('Departure delay, minutes') + 
  ylab('Arrival delay, minutes') +
  theme_minimal()
```

### Faceting

```{r, fig.width = 8, fig.height = 6}
flights %>% select(carrier, dep_delay, arr_delay, origin) %>% left_join(airlines, by="carrier") %>% filter(carrier == c("AA","UA", "DL")) %>% 
  ggplot() + geom_point(aes(x=dep_delay, y=arr_delay, color=name), alpha=.5) +
  ggtitle('Arrival and departure delays, by Airline') +
  xlab('Departure delay, minutes') + 
  ylab('Arrival delay, minutes') +
  theme_minimal() +
  facet_wrap(~origin)
```

### More complex `ggplot2` graphs


Let's use `mtcars` which is a collection of stats on various makes and models of car. 
```{r}
data(mtcars)
mtcars
```

Simple plot of `hp` and `mpg`, with the `cyl` (cylinders) color-coded. Smoothing function added.
```{r}
ggplot(mtcars) + geom_point(aes(hp, mpg, color = factor(cyl))) + # cyl is numeric in this dataset, 
  # but really is categorical rather than continuous
  # so we'll coerce it to be a factor with factor(cyl)
  xlab('Horsepower') +
  ylab('MPG') +
  geom_smooth(aes(hp, mpg), method = "loess") +
  theme_minimal()
```

Smoothing, but using a grouping variable (`cyl`). 

```{r}

ggplot(mtcars, aes(x=hp, y=mpg, color=factor(cyl))) +
  geom_point() + 
  geom_smooth(method=lm) + 
  theme_minimal()

```


# Interactive plots

There are a lot of choices here, each with advantages and disadvantages. But for brevity, we're going to just look at Plotly.js, built on D3.js. The reason I choose this one is primarily due to the connection with `ggplot2`, and the fact that the R Library is free and open-source. 

## `ggplotly` 

```{r, fig.width = 8, fig.height = 8}
library(plotly)
p1 <- ggplot(mtcars, aes(x=hp, y=mpg, color=factor(cyl))) +
  geom_point() + 
  geom_smooth(method=lm) + theme_minimal()

ggplotly(p1)
```

## Types of plots

### Sankey diagrams

```{r}
library(plotly)
library(rjson)

json_file <- "https://raw.githubusercontent.com/plotly/plotly.js/master/test/image/mocks/sankey_energy.json"
json_data <- fromJSON(paste(readLines(json_file), collapse=""))

p <- plot_ly(
    type = "sankey",
    domain = c(
      x =  c(0,1),
      y =  c(0,1)
    ),
    orientation = "h",
    valueformat = ".0f",
    valuesuffix = "TWh",

    node = list(
      label = json_data$data[[1]]$node$label,
      color = json_data$data[[1]]$node$color,
      pad = 15,
      thickness = 15,
      line = list(
        color = "black",
        width = 0.5
      )
    ),

    link = list(
      source = json_data$data[[1]]$link$source,
      target = json_data$data[[1]]$link$target,
      value =  json_data$data[[1]]$link$value,
      label =  json_data$data[[1]]$link$label
    )
  ) %>% 
  layout(
    title = "Energy forecast for 2050<br>Source: Department of Energy & Climate Change, Tom Counsell via <a href='https://bost.ocks.org/mike/sankey/'>Mike Bostock</a>",
    font = list(
      size = 10
    ),
    height = 850, 
    width = 800,
    xaxis = list(showgrid = F, zeroline = F),
    yaxis = list(showgrid = F, zeroline = F)
)

p


```

