---
title: "Functional Programming"
author: "Aaron C Cochran"
date: "March 22, 2018"
output: 
  html_document:
    toc: true
    toc_depth: 1
    toc_float: true
    number_sections: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(purrr)
library(broom)
```

# What is functional programming?

It's the ability to automate the boring stuff and avoid the problem of copying and pasting code for each iterative step in the analysis process. We could spend a whole class on this topic but this exercise will at least introduce you to the powers of functions in R. 

## An example

Using a mock dataset we can write a bunch of steps to rescale each column to have a range of 0 to 1. This code is almost all written in base R. Can you spot the mistake?  
 
```{r}
df <- tibble::tibble(
  a = rnorm(10),
  b = rnorm(10),
  c = rnorm(10),
  d = rnorm(10)
)

df$a <- (df$a - min(df$a, na.rm = TRUE)) / 
  (max(df$a, na.rm = TRUE) - min(df$a, na.rm = TRUE))
df$b <- (df$b - min(df$b, na.rm = TRUE)) / 
  (max(df$b, na.rm = TRUE) - min(df$a, na.rm = TRUE))
df$c <- (df$c - min(df$c, na.rm = TRUE)) / 
  (max(df$c, na.rm = TRUE) - min(df$c, na.rm = TRUE))
df$d <- (df$d - min(df$d, na.rm = TRUE)) / 
  (max(df$d, na.rm = TRUE) - min(df$d, na.rm = TRUE))
```

This is why copying/pasting is a bad practice. Too many opportunities for errors to be introduced, and the code is still not very readable. Yet each step of the code only has a single input, which means we can definitely simplify it. 


```{r}
df <- tibble::tibble(
  a = rnorm(10),
  b = rnorm(10),
  c = rnorm(10),
  d = rnorm(10)
)

rescale01 <- function(x) {
  rng <- range(x, na.rm = TRUE)
  (x - rng[1]) / (rng[2] - rng[1])
}

rescale01(df)
```

By condensing the code into a function we can easily rescale the entire dataset in half the amount of code, and in a much more readble format. 

# `for` loops

Start with our random number `tibble` and compute the median of each column. 

```{r}
df <- tibble::tibble(
  a = rnorm(10),
  b = rnorm(10),
  c = rnorm(10),
  d = rnorm(10)
)

median(df$a)
median(df$b)
median(df$c)
median(df$d)
```

As before, this method has 4 copies of the same code with a different column in each copy. It may be manageable with 4 columns, but much less so with 40.  So we'll use a `for` loop to fix this. 

```{r}
output <- vector("double", ncol(df)) # 1. The output - create an output that is a vector of doubles
# and is as long as the number of columns in the data.frame

for(i in seq_along(df)) { # sequence - for each step along the df from col(1) to col(n)
  output[[i]] <- median(df[[i]]) # body - what we're doing at each iteration
}

output # display the output you defined. 
```

Some key points here:

1. The _output_: `output <- vector("double", length(x))`. Before running a loop you should allocate sufficient space for the output. This is much faster than growing the output organically using `c()`. 

2. The _sequence_: `i in seq_along(df)`. This determines what to loop over. Each run of the loop will assign `i` to a different value from `seq_along(df)`. The `seq_along()` function is a safer version of the base R `1:length(l)`; it is safer because it will give you the right answer with a 0 length vector. 

```{r}
y <- vector("double", 0) # 0 length vector y
1:length(y) # wrong! This is not a length 2 vector with a 1 and a 0

seq_along(y) # correct! This is a vector of interger (doubles) of length 0. 
```

3. The _body_: `output[[i]] <- median(df[[i]])`. This is the workhorse in the function. The first iteration will run `output[[1]] <- median(df[[1]])` and the second iteration will run `output[[2]] <- median(df[[2]])`, etc. 

## Combining `for` loops into a function

Deeper down rabbit hole we go. We can take our `for` loop that calculates median and make it into a function that has more flexibility and can be assigned a name and called on demand. What if we want to be able to calculate a few summary stats from the same function?



```{r}
col_summary <- function(df, fun) { # note this function takes 2 arguments, not one. 
  out <- vector("double", length(df))
  for (i in seq_along(df)) {
    out[i] <- fun(df[[i]])
  }
  out
}


col_summary(df, median)

col_summary(df, mean)

```
This sort of work is so common in R that there is a nice package that does most of it for you. Still, understanding the fundamental code underneath is essential to really understanding how the `purrr` package's `map()` functions work.

# `apply`, and friends
In the base R data wra
# `purrr` 

This package enhances R's functional programming toolkit. The example below is a real-world example of what may be used in a workflow where you are testing the predictive power of a linear model across different segments of a dataset. It uses the `map()` function which maps another function across different columns in a `tbl_df` or `data.frame`. 

```{r}
map_dbl(df, mean)
map_dbl(df, median)
map_dbl(df, sd)

```

We can pipe the `map()` functions as well, making the code even more readable.

```{r}
df %>% map_dbl(mean)

df %>% map_dbl(median)

df %>% map_dbl(sd)

```

How is this better than our `col_summary` loop we wrote?

1. `purrr` functions are implemented in C, making them faster. 
2. The second argument, `.f`, the function to apply, can be a formula, a character vector, or an integer vector. 
3. Additional arugments can be passed to `.f` each time `map()` is called. 
4. The `map()` functions also preserve names. 

```{r}
z <- list(x = 1:3, y=4:5)
map_int(z, length)

```


```{r}
mtcars %>% 
  split(.$cyl) %>% # base R function to split the dataframe by cylinder
  map(~ lm(mpg ~ wt, data=.)) %>% # map a linear model of MPG ~ WEIGHT over each part of the dataset
  map(summary) %>% # summarise each part of the dataset
  map_dbl("r.squared") # extract the R^2 values for the linear model
```

